### 2025-04-08 Paper.tex
Update Paper.tex

>>>>

	\subsubsection{Vocal Extraction}
 Music source separation consists in isolating single sound sources from a mixed audio signal. When the vocal track is extracted from a mix, this practice is usually called vocal extraction. \cite{Vincent2006} categorize source separation into techniques that rely on statistical assumptions (e.g., Independent Component Analysis) and the recent  models that rely on deep learning approaches. In the case of vocal extraction, the methods employed make use of time-frequency representations such as spectrogram masks or raw waveform modeling, with the aim to isolate the singer’s voice while minimizing artifacts and instrumental interferences \cite{Huang2014} Nevertheless, contemporary deep learning methods outperform traditional statistical approaches precisely because they do not operate under the assumptions of independence or basic spectral models. However, this assumption almost never holds true in real-world music recordings, as there is no independence between sources. Conversely, deep learning methods capture complex, nonlinear representations of music directly from data \cite{Uhlich2017}. By learning higher-level features (timbre and harmonic relationships) they are better able to adapt even to recordings, handling as a result more complex scenarios and producing cleaner vocal isolations. The separation process is considerably more robust than with statistical models that tend to falter when sources share similar frequency bands or show correlated dynamics \cite{Vincent2006}.

	The deep neural network approach used for extracting vocals or instruments from a mix can be broken down in two main strategies. The first one is the spectrogram-based separation which converts the audio waveform into a time-frequency representation (usually through the Short-Time Fourier Transform) and subsequently trains a neural network to predict masks or direct magnitudes for each source \cite{Stoeter202} Thus the aim of the spectrogram-based approach is to minimize cross-talk between instruments in the final extracted signal by leveraging both magnitude and phase information in each frequency bin. As a result, each audio frame is split into small frequency bins, each corresponding to a discrete range of frequencies. On the other hand, the second strategy is based on waveform-based separation in the time domain. For instance, Demucs introduced an encoder-decoder architecture operating on raw waveforms that preserves amplitude envelopes more effectively than frequency-domain approaches (Défossez et al., 2019). Yet, while spectrogram-based methods have proven more suitable for certain genres and recordings, time-domain networks offer a superior performance when dealing with fast transients (e.g., drum hits) or complex stereo imaging.

	In our project, we experimented with four different vocal extraction algorithms, namely Demucs, Open-Unmix (UMX), Spleeter, and Ultimate Vocal Remover (UVR) in order to determine the most effective solution for isolating vocals with minimal noise or other instrumental interferences. Each of them leverages deep learning in different ways and is constituted by different architectures to tackle source separation. 
Developed by the French music streaming service Deezer, Spleeter offers multi-stem separation options (2-stem, 4-stem, 5-stem) and operates faster than many alternatives \cite{Stoeter2020}. While Spleeter is highly accessible for quick extractions, its output contains audible cross-talk if the input mix is dense or heavily compressed. The algorithm we fiddled around with was Demucs. Published by Facebook AI Research, it takes a time-domain approach and processes audio directly as a waveform and uses a deep convolutional encoder-decoder network. As a result, it is able to maintain more natural-sounding dynamics. However, just as Défossez et al., 2019 observed, Demucs leaves faint reverb tails or minor bleed-through in the vocal stem. Subsequently, we explored Open-Unmix (UMX), a spectrogram-based method that applies U-Net structures to time-frequency representations of audio signal. UMX systematically masks out non-vocal components in the frequency domain which yields clean vocal extractions \cite{Stoeter2018}. However, in spite of its advantages, UMX tends to introduce phase-related artifacts and the so-called “hollow” tonalities in more complex tracks. Finally, we experimented with Ultimate Vocal Remover (UVR), which uses multiple separation backends including variations of MDX-Net, Demucs, and VR Architecture. Different models can be tested through either a GUI or the command line, picking the algorithm that best suits the purpose. After several thorough comparisons, we came to the conclusion that UVR consistently produced the cleanest vocal stems with barely perceptible instrument bleed or noise, especially when using the latest MDX-based models.  Indeed, its performance can certainly be attributed to the combination of a vast array of available models.
